\pkg{dataMaid}: your assistant for documenting supervised data quality screening in \proglang{R}

### 2021-10-07 docs/CSP2018/slides/module4/module4.Rmd
Update to keep on CRAN

---
title: "Reproducible data screening and quality control"
subtitle: "Module 3: Time to clean and other options"
author: "Claus Thorn Ekstrøm and Anne Helby Petersen<br>UCPH Biostatistics"
date: "eRum, Budapest, May 14th, 2018<br>.footnotesize[[Slides/homepage](https://ekstroem.github.io/dataMaid/eRum2018/index.html)]"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ['default', 'ku.css', 'animate.min.css']
    seal: true
    nature:
      ratio: '16:9'
      slideNumberFormat: '%current%' 
      highlightStyle: ir-black
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "js/zoom.js"
---

@@

.center[
```{r cleanpic, echo=FALSE, out.width="60%"}
knitr::include_graphics("pics/datacleaning.jpg")
```
]

Not the best term ... and should not be unsupervised

@@

Two systems for selecting observations in `data.frame`s in R:

By index (row number) or using a logical vector.

@@

background-image: url(pics/structure.png)
background-size: 30%
background-position: right

## Finishing up after cleaning

@@

.footnotesize[
```{r eval=FALSE}
makeCodebook(presidentData)
```
]

@@

.footnotesize[
```{r}
pD <- presidentData
attr(pD$presidencyYears, "label") <- 
  "Full years as president"
```
]

.footnotesize[
```{r}
attr(pD$birthday, "shortDescription") <- 
  "Dates are stored in YYYY-MM-DD format"
```
]

@@

# Exercise 3

@@

##Row-wise or column-wise checks?

```{r colrow1, echo=FALSE, out.width="75%", fig.align = "center"}
knitr::include_graphics("pics/colrow1.png")
```

##Row-wise or column-wise checks?

```{r colrow2, echo=FALSE, out.width="75%", fig.align = "center"}
knitr::include_graphics("pics/colrow2.png")
```

##Row-wise or column-wise checks?

```{r colrow3, echo=FALSE, out.width="75%", fig.align = "center"}
knitr::include_graphics("pics/colrow3.png")
```

##Row-wise *and* column-wise constraints!

* `dataMaid` performs class dependent checks for each variable in a dataset, one at a time (column-wise)
  + Pros: Easy to document what was (not) done, let's you get started without a lot of prior knowledge, easy to share with collaborators
  + Shortcomings: Generally cannot detect internal consistency issues or use non-class dependent variable constraints

##Row-wise *and* column-wise constraints!

An R-packages that performs row-wise checks: `validate`. Check out the talk on Wednesday @ 14.50 by Edwin de Jonge:

**validatetools - resolve and simplify contradictive or redundant data validation rules**

Note: Different use of the term "validation" - no longer about format, type and range, but used as synonym to "check".

>>>>

---
title: "Cleaning Up the Data Cleaning Process"
subtitle: "Module 4: Time to clean!"
author: "Claus Thorn Ekstrøm and Anne Helby Petersen<br>UCPH Biostatistics"
date: "CSP, Portland, OR, Feb. 15th, 2018<br>.footnotesize[[Slides/homepage](https://ekstroem.github.io/dataMaid/CSP2018/index.html)]"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ['default', 'ku.css', 'animate.min.css']
    seal: true
    nature:
      ratio: '16:9'
      slideNumberFormat: '%current%' 
      highlightStyle: ir-black
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "js/zoom.js"
---

@@

```{r cleanpic, echo=FALSE, out.width="60%"}
knitr::include_graphics("pics/datacleaning.jpg")
```

.pull-right[Not the best term ... and should not be unsupervised]

@@

Two systems for selecting observations in `data.frame`s in R: 
By index (row number) or using a logical vector.

@@

# Exercise 4

background-image: url(pics/structure.png)
background-size: 30%

## Finishing up

@@

.footnotesize[
```{r eval=FALSE}
makeCodebook(bigPresidentData)
```
]

@@

.footnotesize[
```{r}
bPD <- bigPresidentData
attr(bPD$presidencyYears, "label") <- 
  "Full years as president"
```
]

.footnotesize[
```{r}
attr(bPD$dateOfDeath, "shortDescription") <- 
  "Missing means that the person is still alive"
```
]

@@

# Exercise 4b

@@

### 2021-10-07 docs/articles/dataMaid_artData.Rmd
Update to keep on CRAN

---
dataMaid: yes
title: artData
subtitle: "Autogenerated data summary from dataMaid"
date: 2018-02-09 14:33:06
output: html_document
---
```{r , echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(ggplot2)
library(pander)
```
```{r visualFunctions, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
ggAggHist <- getFromNamespace("ggAggHist", "dataMaid")
ggAggBarplot <- getFromNamespace("ggAggBarplot", "dataMaid")
```
# Data report overview
The dataset examined has the following dimensions:

@@

 *  Report creation time: Fri Feb 09 2018 14:33:06

>>>>

---
dataMaid: yes
title: artData
subtitle: "Autogenerated data summary from dataMaid"
date: 2018-02-10 15:57:23
output: html_document
---
```{r , echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
library(ggplot2)
library(pander)
```
```{r visualFunctions, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE, error=FALSE}
ggAggHist <- getFromNamespace("ggAggHist", "dataMaid")
ggAggBarplot <- getFromNamespace("ggAggBarplot", "dataMaid")
```
# Data report overview
The dataset examined has the following dimensions:

@@

 *  Report creation time: Sat Feb 10 2018 15:57:23

### 2021-10-07 latex/article_accept.tex
Update to keep on CRAN

By default, an \proglang{R} markdown file and a rendered pdf,  word or html
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title ``Data
  report overview''. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
``Variable list'': A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the \code{numeric}-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

@@

\smallskip Control summarize, visualize, and check steps \\
\quad \code{summaries} & What summaries should be performed for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{visuals} & What type of visualization should be provided for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{checks} & What checks should be applied to each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities ``summarize'',
                 ``visualize'', ``check'')? &
                                                        \code{c("summarize", "visualize", "check")}  \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{treatXasY} & How should non-supported variable classes be handled? & \code{NULL} (no handling) \\
%Takes a list on the format \code{list(X = Y)} where \code{X} is a non-supported variable class (e.g. \code{complex}) and \code{Y} is a supported one (e.g. \code{numeric}) & \code{NULL} (No rules for handling non-supported variable types) \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{NULL} (use pdf if \LaTeX\ is found, otherwise Word (if on Windows), or html)\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\
\quad \code{vol} & Add a suffix to the file name of the outputted report &   \code{""} (no suffix)\\

@@

 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & \blue{$\times$} & & \blue{$\times$} & \blue{$\times$} \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed white space &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{isCPR} & Identify Danish personal identification numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
 \quad \code{isKey} & Check if the variable is a key & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \hline
\end{tabular}
\caption{Overview of all summary-, visual- and  check functions currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively. The default settings of \code{makeDataReport()} are marked in blue.}
\label{table.SVCfunctions}
\end{table}

@@

Thus we can see, for example, that for \code{numeric}, \code{integer},
and \code{Date} variables, \pkg{dataMaid} provides functions for
adding summary information about the minimum and maximum values, while
all seven variable classes dealt with in \pkg{dataMaid} have functions
for central tendency summaries (i.e., mode or median).

@@

One might also encounter datasets with variables that are not among the 7 classes mentioned in the above (\code{character}, \code{Date}, \code{factor}, \code{integer}, \code{labelled}, \code{logical} and \code{numeric}), for instance variables of type \code{complex} or user-defined classes. \code{makeDataReport()} can be told how to handle such variables by use of the argument \code{treatXasY}. This argument takes a list where the names correspond to "new" variable types (\code{X}), while the entries must be supported variable types (\code{Y}). For instance, we can instruct \code{dataMaid} to treat complex variables as numerics and generate a data report for a type \code{complex} variable like this:

@@

In this report, we will find that the two variables, \code{complexVar} and \code{numericVar} will be have identical presentations in the variable list, as treating a \code{complex} variable as a \code{numeric} means dropping the imaginary part of the complex numbers which was the only thing setting the two variables apart in the first place.

@@

The syntax in this code chunk can be read as follows: "Summarize \code{events} in \code{toyData}, and for \code{numeric} variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=1]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_presidentData_appB.pdf}}\\
\end{center}
\caption{The front page and the first page of the data overview report for the \code{presidentData} dataset. Note that the report title has been customized (front page), \code{identifyLoners} has been removed from the checks performed on character variables ("Identify levels with <6 obs." is not checked for character variables in the table on page 1) and that variables of class \code{Name} have been set to be treated like \code{character} variables (page 1). Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP01}
\end{figure}

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=4]{dataMaid_presidentData_appB.pdf}}
\end{center}
\caption{The second and third pages of the \code{presidentData} data report. We see that there are two \code{Name} variables in the overview on page 2 and see that these variables are indeed treated as \code{character} variables on page 3, as specified in the \code{makeDataReport} call by use of the \code{treatXasY} argument. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP23}
\end{figure}

@@

We discuss the results of a data overview report generated for this dataset below, but first there are a few special features of the dataset and wishes for the data report that require us to customize it using some of the arguments for \code{makeDataReport}. We have the following points of interest:
\begin{enumerate}
\item A couple of variables are used to store names, namely \code{lastName} and \code{firstName}. In order to use special bibliographical analysis tools on these, and only these, variables, it might be convenient to assign them a special class. Therefore, these variables have been set to have class \code{Name} by use of the base \proglang{R} function \code{class()}. When we wish to make a data report for the dataset, we have to tell \code{makeDataReport()} how to handle such \code{Name}-type variables, as they are not among the supported variable types mentioned in the above. This can be done using the \code{treatXasY} argument.
\item We use the \code{character} class for a few categorical variables that have a lot of different levels, but where it is not a data mistake that these levels each generally have very few observations, e.g., the variable \code{stateOfBirth}. Therefore, we would like to disable the \code{identifyLoners} check (which flags variables with <6 observations in any of the levels) for this variable type. This can be done using the \code{checks} argument.
\item We would like the report to be called "Dirty president data" in order to reflect that it is, indeed, a report concerning dirty data about presidents.
\end{enumerate}
Incorporating these three customizations, we can load the data and generate a report by calling:

@@

On page 3, we see the contents for the three first variables. Here, we identify a prefixed white space in the last name entry for President Truman and we find that a dot was entered as a first name; this is a typical choice for coding missing values in e.g., \proglang{Stata}, and therefore, it is flagged as a potential miscoded missing value. The variable \code{orderOfPresidency} is not summarized, visualized or checked because it is categorical and contains unique values for each observation.

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the \code{birthday} variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an infinite presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As \code{complex} is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does not reflect a mistake in the data, but rather a peculiarity in the real world, and as such, it should not be cleaned out.

@@

Please note that if \code{ageAtInauguration} had been a \code{factor} rather than a \code{character} variable, an additional inner call should be added in order to ensure no conversion issues:

@@

Finally, we create a new data report, adding the suffix "cleaned" to the title, as well as the file name, so that we have documentation of the current state of the dataset. We also decide that it might be sufficient to be able to inspect only the real part of the presidential favorite numbers and therefore, we choose to treat the \code{complex} variable \code{favoriteNumber} as a \code{numeric} variable:

@@

\subsection[Incorporating dataMaid in an automated workflow]{Incorporating \pkg{dataMaid} in automated workflow}
The default settings of \code{makeDataReport()} have been set to facilitate easy and quick data report generation, but unfortunately, this also means that it is not ideal for a more programming-oriented workflow, where the function might not be called by a human. For instance, one might be interested in automatically running \code{makeDataReport()} on all datasets received from a certain client, perhaps via email or through a web upload, and returning a data report for them to inspect and comment before ever looking at the data. In this scenario, there are a few issues with the standard data report:
\begin{enumerate}
\item After rendering, the report is automatically opened. This is not very useful, if the processes are supposed to run in the background.
\item The report generation writes messages to the console while producing and rendering the report.
\item Unless specifically told otherwise, every report created for the same dataset (or different datasets with the same storage name in \proglang{R}) will have the same file name.
\end{enumerate}
Please note that the data report does contain information about who, when and how concerning its generation, so even though the default choices for file names do not make it easy to tell different reports for the same dataset apart, it should be rather easy when inspecting the report manually.

@@

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output  can be obtained directly in the console by using the \code{check()} function interactively. When called on a \code{data.frame}, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unraveled using the list manipulation function
\code{sapply()}:

@@

{\small
\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

```{r, child = 'dataMaid_toyData.Rmd'}
```
\end{Verbatim}
}

In the this example, the \code{dataMaid\_toyData.Rmd} file could have
been created as follows:

@@

Note that with the latter option, no special YAML preamble is needed in the \proglang{R} markdown document.

Alternatively, one can create the usual output report, not render it and then manually edit the produced \proglang{R} markdown file as wished. This can be done with the following command:

>>>>

By default, an \proglang{R} markdown file and a rendered pdf,  word or html
overview document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three
parts, two of which are presented in Figure~\ref{fig:example1}. First,
an overview of what was done is presented under the title ``Data
  report overview''. Secondly, an index listing each variable along
with an indication of whether it was found to be problematic or not is
provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable
in the dataset is presented in turn using (up to) three tools in the
``Variable list'': A table summarizing key features of the
variable, a figure visualizing its distribution when relevant and a list of flagged
issues, if any. For instance, as shown in Figure~\ref{fig:example2},
for the numeric-type variable \code{events} from
\code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and
\code{NaN}), while two values were also flagged as potential outliers
that should be investigated more carefully.

@@

\smallskip Control summarize, visualize, and check steps \\
\quad \code{summaries} & What summaries should be performed for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{visuals} & What type of visualization should be provided for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{checks} & What checks should be applied to each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities ``summarize'',
                 ``visualize'', ``check'')? &
                                                        \code{c("summarize", "visualize", "check")}  \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{treatXasY} & How should non-supported variable classes be handled? & \code{NULL} (no handling) \\
%Takes a list on the format \code{list(X = Y)} where \code{X} is a non-supported variable class (e.g. complex) and \code{Y} is a supported one (e.g. numeric) & \code{NULL} (No rules for handling non-supported variable types) \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{NULL} (use pdf if \LaTeX\ is found, otherwise Word (if on Windows), or html)\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \proglang{R} markdown file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\
\quad \code{vol} & Add a suffix to the file name of the outputted report &   \code{""} (no suffix)\\

@@

 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & \blue{$\times$} & & \blue{$\times$} & \blue{$\times$} \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed white space &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{isCPR} & Identify Danish personal identification numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
 \quad \code{isKey} & Check if the variable is a key & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \hline
\end{tabular}
\caption{Overview of all summary-, visual- and  check functions currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to character, factor,
  integer, labelled, logical (boolean), numeric, and Date, respectively. The default settings of \code{makeDataReport()} are marked in blue.}
\label{table.SVCfunctions}
\end{table}

@@

Thus we can see, for example, that for numeric, integer,
and Date variables, \pkg{dataMaid} provides functions for
adding summary information about the minimum and maximum values, while
all seven variable classes dealt with in \pkg{dataMaid} have functions
for central tendency summaries (i.e., mode or median).

@@

One might also encounter datasets with variables that are not among the 7 classes mentioned in the above (character, Date, factor, integer, labelled, logical and numeric), for instance variables of type complex or user-defined classes. \code{makeDataReport()} can be told how to handle such variables by use of the argument \code{treatXasY}. This argument takes a list where the names correspond to "new" variable types (\code{X}), while the entries must be supported variable types (\code{Y}). For instance, we can instruct \code{dataMaid} to treat complex variables as numeric and generate a data report for a type complex variable like this:

@@

In this report, we will find that the two variables, \code{complexVar} and \code{numericVar} will be have identical presentations in the variable list, as treating a complex variable as a numeric means dropping the imaginary part of the complex numbers which was the only thing setting the two variables apart in the first place.

@@

The syntax in this code chunk can be read as follows: "Summarize \code{events} in \code{toyData}, and for numeric variables, set the summaries to be the default summary functions, except \code{variableType} and \code{countMissing}."

@@

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=1]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_presidentData_appB.pdf}}\\
\end{center}
\caption{The front page and the first page of the data overview report for the \code{presidentData} dataset. Note that the report title has been customized (front page), \code{identifyLoners} has been removed from the checks performed on character variables ("Identify levels with <6 obs." is not checked for character variables in the table on page 1) and that variables of class \code{Name} have been set to be treated like character variables (page 1). Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP01}
\end{figure}

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_presidentData_appB.pdf}}
\frame{\includegraphics[width=7.5cm,page=4]{dataMaid_presidentData_appB.pdf}}
\end{center}
\caption{The second and third pages of the \code{presidentData} data report. We see that there are two \code{Name} variables in the overview on page 2 and see that these variables are indeed treated as character variables on page 3, as specified in the \code{makeDataReport} call by use of the \code{treatXasY} argument. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix2}.}
\label{fig:bigExampleP23}
\end{figure}

@@

We discuss the results of a data overview report generated for this dataset below, but first there are a few special features of the dataset and wishes for the data report that require us to customize it using some of the arguments for \code{makeDataReport}. We have the following points of interest:
\begin{enumerate}
\item A couple of variables are used to store names, namely \code{lastName} and \code{firstName}. In order to use special bibliographical analysis tools on these, and only these, variables, it might be convenient to assign them a special class. Therefore, these variables have been set to have class \code{Name} by use of the base \proglang{R} function \code{class()}. When we wish to make a data report for the dataset, we have to tell \code{makeDataReport()} how to handle such \code{Name}-type variables, as they are not among the supported variable types mentioned in the above. This can be done using the \code{treatXasY} argument.
\item We use the character class for a few categorical variables that have a lot of different levels, but where it is not a data mistake that these levels each generally have very few observations, e.g., the variable \code{stateOfBirth}. Therefore, we would like to disable the \code{identifyLoners} check (which flags variables with <6 observations in any of the levels) for this variable type. This can be done using the \code{checks} argument.
\item We would like the report to be called "Dirty president data" in order to reflect that it is, indeed, a report concerning dirty data about presidents.
\end{enumerate}
We incorporate these three customizations, load the data, and generate a report by calling:

@@

On page 3, we see the contents for the three first variables. Here, we identify a prefixed white space in the last name entry for President Truman and we find that a dot was entered as a first name; this is a typical choice for coding missing values in e.g., \proglang{Stata} \citep{stata}, and therefore, it is flagged as a potential miscoded missing value. The variable \code{orderOfPresidency} is not summarized, visualized or checked because it is categorical and contains unique values for each observation.

Figure \ref{fig:bigExampleP45} presents the remaining two pages with variable presentations. On pages 4 and 5, we find a few remarks:
\begin{itemize}
\item In the \code{birthday} variable, there is an entry with the date March 1, 1300 which is a bit of an outlier. 
\item Among the states in which the presidents were born, New York was mistakenly spelled with a lower case "Y" in at least one entry.
\item The variable concerning assassination attempts is coded as a numeric, but the default \code{smartNum = TRUE} setting of \code{makeDataReport()} implies that such a numeric variable with only a few (less that 5) unique values is treated as a factor variable in the data report, thereby providing more relevant summaries, visualizations and checks. This is remarked in the variable presentation for \code{assassinationAttempt} and it can also be seen by the visualization being a barplot rather than a histogram.
\item The variable concerning the sex of the president was skipped, as there is nothing to present when all US presidents have been male so far.
\item The report flags the variable \code{ethnicity} to have suspiciously few observations in one category, "African American". 
\item A few presidents were found to have odd values in the variable describing the duration of their presidencies: Some had very short (outlier) presidencies of less than two years and one was registered to have an infinite presidency. 
\item The variable concerning age at inauguration was coded as a character variable, but consists exclusively of numbers and takes a lot of different values and therefore, it was flagged as a potentially misclassified numeric variable.
\item It seems as if one or more presidents have complex numbers as their favorite numbers. As complex is not a supported variable type in \code{dataMaid} and no strategy for handling this class was provided in the \code{treatXasY} argument, the variable is simply flagged as non-supported.
\end{itemize}
A lot of these mistakes are easily fixable, and we will do so below. However, some of them require more delicate knowledge of the subject matter. For instance, \code{ethnicity} is very reasonably marked as a potentially problematic variable as it includes only a single observation of "African American". However, a human reading this report will know that this does not reflect a mistake in the data, but rather a peculiarity in the real world, and as such, it should not be cleaned out.

@@

Please note that if \code{ageAtInauguration} had been a factor rather than a character variable, an additional inner call should be added in order to ensure no conversion issues:

@@

Finally, we create a new data report, adding the suffix "cleaned" to the title, as well as the file name, so that we have documentation of the current state of the dataset. We also decide that it might be sufficient to inspect only the real part of the presidential favorite numbers and therefore, we choose to treat the complex variable \code{favoriteNumber} as a numeric variable:

@@

\subsection[Incorporating dataMaid in an automated workflow]{Incorporating \pkg{dataMaid} in automated workflow}
The default settings of \code{makeDataReport()} have been set to facilitate easy and quick data report generation, but unfortunately, this also means that it is not ideal for a more programming-oriented workflow, where the function might not be called by a human. For instance, one might be interested in automatically running \code{makeDataReport()} on all datasets received from a certain client, perhaps via email or through a web upload, and returning a data report for them to inspect and comment before ever looking at the data. In this scenario, there are a few issues with the standard data report:
\begin{enumerate}
\item After rendering, the report is automatically opened. This is not very useful, if the processes are supposed to run in the background.
\item The report generation writes messages to the console while producing and rendering the report.
\item Unless specifically told otherwise, every report created for the same dataset (or different datasets with the same storage name in \proglang{R}) will have the same file name.
\end{enumerate}
Note that the data report does contain information about who, when and how concerning its generation, so even though the default choices for file names do not make it easy to tell different reports for the same dataset apart, it should be rather easy when inspecting the report manually.

@@

Now only the checking results are printed, and only for variables
where problems were identified. An even more minimal output  can be obtained directly in the console by using the \code{check()} function interactively. When called on a data frame, this function produces a list (of
variables) of lists (of checks) of lists (or rather,
\code{checkResult}s). Thus, the overall problem status of each variable
can easily be unraveled using the list manipulation function
\code{sapply()}:

@@

\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\rule{\textwidth}{0.4pt}}
  - \newcommand{\bminione}{\begin{minipage}{0.75\textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25\textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

```{r, child = 'dataMaid_toyData.Rmd'}
```
\end{Verbatim}

In this example, the \code{dataMaid\_toyData.Rmd} file could have
been created as follows:

@@

Note that with the latter option, no special YAML preamble is needed in the \proglang{R} markdown document. 
%
Alternatively, one can create the usual output report, not render it and then manually edit the produced \proglang{R} markdown file as wished. The following command does that:
